### 2021-07-10 docs/manuscript/manuscript.tex
Update manuscript.tex

In this work, we have conducted extensive computational explorations of small
GRNs with the capacity to generate a simple spatial pattern in a 1-D
morphogenetic field. Our findings shed new light on regulatory rules and
dynamical mechanisms that  GRNs observed in nature could potentially
implement to achieve fundamental spatio-temporal tasks, which might
serve as guidance for future synthetic biology implementations.

Overall, based on our extensive search space exploration we were able to identify
a great variety of pattern-forming GRN topologies  with distinctive and often
widely shared building blocks (network motifs), which are likely to confer on
each topology the ability to achieve specific regulatory tasks in a semi-autonomous
manner. However, we also found a great diversity of network
topologies that could produce a striped gene expression pattern, where 615 out of
714 were multi-input topologies, showing that considering multiple
morphogen inputs allows to find different solution for the patterning
problem. Most importantly, we found that this ensemble of pattern-forming GRNs
tend to form a highly connected meta-graph, which highlights at a very high level
emergent properties, such as robustness and evolvability, of complex biological
systems \cite{kitano2004}.\\

Robustness is a defining feature of any developmental patterning process,
and our results suggests that the underlying morphogen interpretation
mechanisms (GRNs) are highly accessible through exploration of the
topology space constrained not only by selection for their final steady state,
but also by their ability to buffer noise~\cite{lo_robust_2015, exelby_2021}.
In our study, the most abundant topologies were found forming highly
connected clusters in a so-called neutral network of genotypes,
implying that the corresponding GRN topologies would tend to be robust
to quantitative changes in the strength of regulatory interactions, diffusion and
degradation parameters, as well as to qualitative changes in the topology
itself. In essence, such organization could be seen as an intrinsic evolutionary
property of GRNs that facilitates the coexistence of robustness and
evolvability~\cite{wagner_robustness_2008}. For instance, bridge topologies
in the neutral network connect distinct classes of pattern-forming
GRNs~(\nameref{S6_Fig}), which suggests that an organism could switch
between different GRNs (for example between a Gierer-Meinhardt-like network
to an Incoherent Feed-Forward Loop) with minimal fitness costs. This high-level
property of the neutral network of topologies could thus facilitate the fine-tuning
of the regulatory systems through genetic modifications in order to meet
specific functional requirements under selective pressures.

In contrast to this robustness to
qualitative and quantitative changes, we found that many GRNs were susceptible
to changes in morphogenetic field size, but interestingly, GRNs with the same
topology behave in different ways for each field size (\nameref{S3_Fig}C),
indicating that the specific values of network parameters are more important
than the network topology when it comes to producing the same pattern at
different scales.

Other studies have shown that Incoherent Feed Forward Loops are able to form
spatial stripes of gene expression~\cite{ishihara_cross_2005}. In particular,
we found that the I3-FFL is one of the network motifs that appears more
frequently in these GRNs, as has also been reported by Cotterell \&
Sharpe~\cite{Cotterell2010}, and has been found in transcriptional networks involved
in cell fate definition~\cite{Li2019}. Moreover, our results indicate that this
is a robust motif that can be useful for construction of synthetic gene
regulatory networks that are exposed to stochastic fluctuations in the
environment and mutations.

However, as our model was less stringent and allowed the morphogen input to
interact with any gene, we found a larger number of possible topologies with the
same striped phenotype and a diverse combination of diffusion and degradation
parameters (\nameref{S7_Fig}).
For example, in a single input model the I4-FFL could
not produce a stripe of gene expression~\cite{munteanu_2014}, whereas in our
experimental setup 20.17\% of the topologies presented this motif as a
subgraph~(\nameref{S1_Table}).

Although it has been experimentally demonstrated that the Incoherent Feed
Forward Loop type 2 (I2-FFL) is able to generate a striped gene expression
pattern,~\cite{Schaerli2014, Basu2005}, in our study this motif was present in
only 10.92\% of the topologies. This fact may be because evolutionary algorithms
can not always find global minima but local minima or its surroundings, however,
it is also possible that the I2-FFL motif is not as robust as other motifs and
therefore be less represented.

@@

The correlation between the complexity index of a topology and its abundance
indicates that networks with simpler designs are more robust to changes in
interaction parameters and that more complex networks are less robusts. For
example, topology number 420 (Fig~\ref{fig:2neut-net}), which
presents high complexity, was one of the
topologies with lower abundance and was disconnected from all other topologies
in the two-step neighbors in the neutral network.

The cladogram of spatiotemporal expression profiles shows that the search space
algorithm selected similar expression mechanisms that can be produced by very
different topologies.

Although GRNs that produce a band of gene expression with only two nodes have
been described (e.g. the I-zero motif,~\cite{Schaerli2014}) this is to our
knowledge the first time a GRN with a topology such as that of topology nine
(Fig \ref{fig:topol9})
and its expression mechanism are reported, increasing our knowledge of the
genotype-phenotype map in this kind of system, and it would be very interesting
to confirm experimentally the ability of this topology to generate the described
pattern and it would be interesting to study its feasibility and possible
implementation in synthetic gene regulatory networks.

This work provides a catalog of GRNs that can be extended in future computational
and theoretical analysis and provides a foundation to design and build synthetic
regulatory networks or to identify new ones existing in nature.

@@

%Aliquam tincidunt, ex in facilisis elementum, libero lectus luctus est, non vulputate nisl augue at dolor. For more information, see \nameref{S1_Appendix}.

@@

\section*{Acknowledgements}
The computational resources (Stevin Supercomputer Infrastructure) and
  services used in this work were provided by the VSC (Flemish Supercomputer
  Center), funded by Ghent University, FWO and the Flemish Government –
  department EWI.

>>>>

In this work, we have conducted extensive computational explorations of small
GRNs with the capacity to generate a simple spatial pattern in a 1-D
morphogenetic field. Our findings shed new light on regulatory rules and
dynamical mechanisms that  GRNs observed in nature could potentially
implement to achieve fundamental spatio-temporal tasks, which might
guide for future synthetic biology implementations.

Overall, based on our extensive search space exploration we identified
a great variety of pattern-forming GRN topologies  with distinctive and often
widely shared building blocks (network motifs), which are likely to confer
each topology the ability to achieve specific regulatory tasks in a semi-autonomous
manner. However, we also found a great diversity of network
topologies that could produce a striped gene expression pattern, where 615 out of
714 were multi-input topologies, showing that considering multiple
morphogen inputs allows finding a different solution for the patterning
problem. Most importantly, we found that this ensemble of pattern-forming GRNs
tends to form a highly connected meta-graph, which highlights emergent
properties, such as robustness and evolvability, of complex biological
systems at a very high level~\cite{kitano2004}.\\

Robustness is a defining feature of any developmental patterning process.
Our results suggest that the underlying morphogen interpretation
mechanisms (GRNs) are highly accessible by exploring the
topology space constrained by selection for their final steady-state and
by their ability to buffer noise~\cite{lo_robust_2015, exelby_2021}.
In our study, the most abundant topologies were found forming highly
connected clusters in a so-called neutral network of genotypes,
implying that the corresponding GRN topologies would tend to be robust
to quantitative changes in the strength of regulatory interactions, diffusion, and
degradation parameters, as well as to qualitative changes in the topology
itself. In essence, such organization could be seen as an intrinsic evolutionary
property of GRNs that facilitates the coexistence of robustness and
evolvability~\cite{wagner_robustness_2008}. For instance, bridge topologies
in the neutral network connect distinct classes of pattern-forming
GRNs~(\nameref{S6_Fig}), suggesting that an organism could switch
between different GRNs (for example, between a Gierer-Meinhardt-like network
to an Incoherent Feed-Forward Loop) with minimal fitness costs. This high-level
property of the neutral network of topologies could thus facilitate the fine-tuning
of the regulatory systems through genetic modifications to meet
specific functional requirements under selective pressures.

In contrast to this robustness to
qualitative and quantitative changes, we found that many GRNs were susceptible
to changes in morphogenetic field size. However, interestingly, GRNs with the same
topology behave in different ways for each field size~(\nameref{S3_Fig}C),
indicating that the specific values of network parameters are more important
than the network topology when it comes to producing the same pattern at
different scales.

Other studies have shown that Incoherent Feed-Forward Loops can form
spatial stripes of gene expression~\cite{ishihara_cross_2005}. In particular,
we found that the I3-FFL is one of the network motifs that appears more
frequently in these GRNs, as has also been reported by Cotterell \&
Sharpe~\cite{Cotterell2010}, and has been found in transcriptional networks involved
in cell fate definition~\cite{Li2019}. Moreover, our results indicate that this
is a robust motif that can be useful for constructing synthetic gene
regulatory networks exposed to stochastic fluctuations in the
environment and mutations.

However, as our model was less stringent and allowed the morphogen input to
interact with any gene, we found a more significant number of possible topologies with the
same striped phenotype and a diverse combination of diffusion and degradation
parameters~(\nameref{S7_Fig}). For example, in a single input model, the I4-FFL could
not produce a stripe of gene expression~\cite{munteanu_2014}, whereas, in our
experimental setup, 20.17\% of the topologies presented this motif as a
subgraph~(\nameref{S1_Table}).

Although it has been experimentally demonstrated that the Incoherent Feed-Forward
Loop type 2 (I2-FFL) can generate a striped gene expression
pattern~\cite{Schaerli2014, Basu2005}, in our study, this motif was present in
only 10.92\% of the topologies. This fact may be because evolutionary algorithms
can not always find global minima but local minima or their surroundings. However,
it is also possible that the I2-FFL motif is not as robust as other motifs and
therefore be less represented.

@@

The correlation between the complexity index of a topology and its abundance
indicates that networks with simpler designs are more robust to changes in
interaction parameters and that more complex networks are less robust. For
example, topology number 420~(Fig~\ref{fig:2neut-net}), which
presents high complexity, was one of the
topologies with lower abundance and was disconnected from all other topologies
in the two-step neighbors in the neutral network.

The cladogram of spatiotemporal expression profiles shows that the search space
algorithm selected similar expression mechanisms that very
different topologies can produce.

Although GRNs that produce a band of gene expression with only two nodes have
been described (e.g., the I-zero motif~\cite{Schaerli2014}), this is to our
knowledge the first time a GRN with a topology such as that of topology
nine~(Fig \ref{fig:topol9})
and its expression mechanism is reported, increasing our knowledge of the
genotype-phenotype map in this kind of system. It would be very interesting
to confirm experimentally the ability of this topology to generate the described
pattern and it would be interesting to study its feasibility and possible
implementation in synthetic gene regulatory networks.

This work provides a catalog of GRNs that can be extended in future computational
and theoretical analysis and provides a foundation to design and build synthetic
regulatory networks or identify new ones existing in nature.

@@

@@

\section*{Acknowledgements}
The computational resources (Stevin Supercomputer Infrastructure) and
services used in this work were provided by the VSC (Flemish Supercomputer
Center), funded by Ghent University, FWO and the Flemish Government–department EWI.
All authors acknowledge partial support from CoDI-Universidad de Antioquia under
project "Estudio de la Evolución de Circuitos de Regulación del Desarollo Embrionario a través de
la Biología Evolutiva de Sistemas" code 2017-14367. All authors acknowledge the Colombian
youth for their defense of a better country.
